On the Gravitational stability of gravito-turbulent accretion disks

### 2016-03-24 paper/model3d.tex
revising issue 8 now

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}\label{ideal_eos}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
{\bf For simplicity we only refer to the adiabatic index as that
  appears in the equation of state for both 2D and 3D disk models 
  \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
                             %for the same 
                             %disk in 2D 
%johnson: specify 3D gamma, then translate into 2D gamma because disk
%model is 2D 
}
The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

For beta cooling given by Eq. \ref{beta_cool} we assume an 
equilibrium polytropic relation 
\begin{align}\label{poly_vert} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(z=0)$ is the equilibrium mid-plane density. {\bf Note that
  $\Gamma$ is only relevant 
  to the 3D problem as it determines the disk vertical structure.} 
The vertical structure is first obtained from 
Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
then inserted into 
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right),
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align} 
{\bf with 
  \begin{align}\label{tirr_def}
    \theta = \frac{\tirr}{T_\mathrm{eq}},
  \end{align}
   where $T_\mathrm{eq}(z)$ is the equilibrium temperature field.
   We shall consider vertically isothermal disks with $\Gamma=1$, as
  appropriate for the outer parts of irradiated protoplanetary disks
   \citep{chiang97}. In this case  
   $\alpha$, $T_\mathrm{eq}$ and $\theta$ are simply constants. 
    Note that $\theta$  
    should only be interpreted as an irradiation parameter when it is 
    defined through Eq. \ref{tirr_def}, in conjunction with adopting  
    Eq. \ref{beta_cool} as the cooling function. 
 }

>>>>

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}\label{ideal_eos}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
{\bf For simplicity we only refer to the adiabatic index as that
  appears in Eq. \ref{ideal_eos} for both 2D and 3D disk models 
  \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
                             %for the same 
                             %disk in 2D 
%johnson: specify 3D gamma, then translate into 2D gamma because disk
%model is 2D 
}
The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

For beta cooling given by Eq. \ref{beta_cool} we assume an 
equilibrium polytropic relation 
\begin{align}\label{poly_vert} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
{\bf where $\rho_0 = \rho(z=0)$ is the equilibrium mid-plane density, and    
$\Gamma$ is the constant polytropic index which determines the disk's
  vertical structure. Thus $\Gamma$ is only relevant to the 3D
  problem.}     
The vertical structure is first obtained from 
Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
then inserted into 
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right),
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align} 
{\bf with 
  \begin{align}\label{tirr_def}
    \theta = \frac{\tirr}{T_\mathrm{eq}},
  \end{align}
   where $T_\mathrm{eq}(z)$ is the equilibrium temperature field.
   We shall consider vertically isothermal disks with $\Gamma=1$, as
  appropriate for the outer parts of irradiated protoplanetary disks
   \citep{chiang97}. In this case  
   $\alpha$, $T_\mathrm{eq}$ and $\theta$ are simply constants. 
    Note that $\theta$  
    should only be interpreted as an irradiation parameter when it is 
    defined through Eq. \ref{tirr_def}, in conjunction with adopting  
    Eq. \ref{beta_cool} as the cooling function. 
 }

### 2016-03-24 paper/results2d.tex
revising issue 8 now

\subsection{Inviscid limit}\label{2d_inviscid}
We first simplify the problem by setting $\alpha = \alpha_b = 0$. This
eliminates viscous heating and forces in the linearized problem, 
allowing us to quantify the \emph{sole} effect of cooling on the 
perturbations. %KMK
We emphasize that destabilization is independent of the effect of decreasing temperature on the instantaneous
value of the classic-$Q$. A time-independent heat source should be
invoked to balance the imposed cooling to allow an equilibrium to be
defined ($\mathcal{H}_\mathrm{ext}\neq 0$). For example, we could assume
that the viscosity only provides a background heating and does not
play an active role in the perturbed state.

@@

\begin{align}\label{betastar}
  \beta_* = \frac{1}{\left(\sqrt{\gamma} - 1\right)^{3/2}}. 
\end{align}
This equation gives remarkably similar values 
of the cooling times below which 
numerical simulations show dynamical disk fragmentation 
\citep{gammie01,rice05,rice11}. These simulations employ the same beta  
cooling prescription with $\theta=0$, and determine the fragmentation
boundary, $\beta_c$, as a function of the adiabatic index $\gamma$.  
Table \ref{bstar_compare} shows rough agreement between $\beta_*$ and 
$\beta_c$.

@@

%KMK moved fig ref. 
To compute growth rates numerically, 
we consider a model with $\gamma=1.4$, $\alpha_b=0$ and 
$\alpha=\alpha(\beta)$ given by thermal equilibrium
(Eq. \ref{alpha_beta_relation}). Furthermore, we relate   
\begin{align}
  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
\end{align}
%with $Q_\mathrm{crit}$ given by 
to mimic a basic state maintained by gravito-turbulence where one
might expect the dimensionless stress $\alpha \sim Q^{-2}$
\citep{lin87}.

@@

%{\bf maybe move following to discussion}
Numerical simulations of gravito-turbulent disks show there is a
maximum $\alpha$ ($\sim 0.06$) that can be sustained before
fragmentation \citep{rice05}. We can interpret this result in our
linear framework. 
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$ (second case in
Eq. \ref{high_visc}). Fig. \ref{gammie_maxrate_plot}  
shows that such a disk would be dynamically unstable with 
growth rate $s = O(\Omega)$. This is due to the direct effect of viscous 
stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
and hence highly turbulent, self-gravitating disks to persist beyond
dynamical timescales.
%KMK
We might interpret fragmentation as the response to the instability of the
highly viscous state. This is consistent with previous numerical
simulations performed by \cite{lodato05}. 
%Case (\ref{gvisc}) is consistent with 
\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    1.0cm]{figures/result2d_gvisc_kmax}
  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
    viscous GI as a function of cooling time for the
    case shown in the top panel of Fig. \ref{gammie_rate_plot}. Black curves 
    computed from the dispersion relation (Eq. \ref{thindisk}), and
    the orange curves are estimates based on
    Eq. \ref{gammie_smallk}---\ref{gammie_bigk}. 
    The dashed
    line marks the region with $\alpha > 1$, and the 
    dashed-dot line marks the region where growth timescales exceed
    the dynamical time.   
    \label{gammie_maxrate_plot}}
\end{figure}

>>>>

\subsection{Inviscid limit}\label{2d_inviscid}
We first simplify the problem by setting $\alpha = \alpha_b = 0$. This
eliminates viscous heating and forces in the linearized problem, 
allowing us to quantify the \emph{sole} effect of cooling on the 
perturbations. %KMK 
We emphasize that destabilization is independent of the effect of
decreasing temperature on the instantaneous value of the
classic-$Q$. A time-independent heat source should be invoked to
balance the imposed cooling to allow an equilibrium to be 
defined ($\mathcal{H}_\mathrm{ext}\neq 0$).

{\bf
For example, we could assume that the viscosity only provides a
background heating and does not play an active role in the perturbed
state. This is in fact done implicitly in the literature when
discussing fragmentation or clump formation in cooling, 
self-gravitating disk simulations \citep{gammie01}. 
There, the effect of the ambient 
gravito-turbulent viscosity on the forming-clump is neglected, as 
one only compares adiabatic heating and the imposed cooling.  
This comparison is encapsulated in the generalized dispersion relation
below.}

@@

\begin{align}\label{betastar}
  \beta_* = \frac{1}{\left(\sqrt{\gamma} - 1\right)^{3/2}}. 
\end{align}
This equation gives similar values 
of the cooling times below which 
numerical simulations show dynamical disk fragmentation 
\citep{gammie01,rice05,rice11}. These simulations employ the same beta  
cooling prescription with $\theta=0$, and determine the fragmentation
boundary, $\beta_c$, as a function of the adiabatic index $\gamma$.  
Table \ref{bstar_compare} shows rough agreement between $\beta_*$ and 
$\beta_c$.  {\bf The match is remarkable, especially with the
  global 3D simulations of \cite{rice05}, since Eq. \ref{betastar}
  is derived for 2D disks in the local limit.}

@@

%KMK moved fig ref. 
To compute growth rates numerically, 
we consider a model with $\gamma=1.4$, $\alpha_b=0$ and 
$\alpha=\alpha(\beta)$ given by thermal equilibrium
(Eq. \ref{alpha_beta_relation}). Furthermore, we relate
{\bf the strength of self-gravity and viscosity by}   
\begin{align} 
  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
\end{align}
%with $Q_\mathrm{crit}$ given by 
{\bf to mimic a gravito-turbulent basic state,} where one
might expect the dimensionless stress $\alpha \sim Q^{-2}$
\citep{lin87}.

@@

%{\bf maybe move following to discussion}
Numerical simulations of gravito-turbulent disks show there is a
maximum $\alpha$ ($\sim 0.06$) that can be sustained before
fragmentation \citep{rice05}. We can interpret this result in our
linear framework. 
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$ (second case in
Eq. \ref{high_visc}). Fig. \ref{gammie_maxrate_plot}  
shows that such a disk would be dynamically unstable with 
growth rate $s = O(\Omega)$. This is due to the direct effect of viscous 
stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
and hence highly turbulent, self-gravitating disks to persist beyond
dynamical timescales.
%KMK
We might interpret fragmentation {\bf as a reflection of} instability
of the highly viscous state. This is consistent with previous numerical
simulations performed by \cite{lodato05}. 
%Case (\ref{gvisc}) is consistent with 
\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    1.0cm]{figures/result2d_gvisc_kmax}
  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
    viscous GI as a function of cooling time for the
    case shown in the top panel of Fig. \ref{gammie_rate_plot}. Black curves 
    computed from the dispersion relation (Eq. \ref{thindisk}), and
    the orange curves are estimates based on
    Eq. \ref{gammie_smallk}---\ref{gammie_bigk}. 
    The dashed
    line marks the region with $\alpha > 1$, and the 
    dashed-dot line marks the region where growth timescales exceed
    the dynamical time.   
    \label{gammie_maxrate_plot}}
\end{figure}
